Multilevel generalizations of collective spin dynamics and squeezing

### 2019-10-25 collective_sun/multilevel_spin_notes.tex
minor changes to math and text

For our final re-expression of collective SU($n$)-symmetric
interactions, we abandon the use of single-spin operators entirely to
express the interaction Hamiltonian $H_{\t{int}}$ in terms of
operators that permute tensor factors (subsystems) of a Hilbert space
with an $N$-fold tensor product structure (e.g.~an array of nuclear
spins).  Specifically, we denote the Hilbert space of an $n$-level
system by $\H_n$, denote its $N$-fold tensor product by
$\H_n^{\otimes N}$, and identify the set $\Pi\equiv\set{\Pi_\sigma}$
of operators that permute tensor factors of $\H_n^{\otimes N}$
according to elements $\sigma\in\SS_N$ of the symmetric group of order
$N$.  We then expand
\begin{align}
  H_{\t{int}}
  = -\f{u}{2} \sum_{p,q,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  = -u\sum_{\substack{p<q\\\mu,\nu}} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  - \f{u}{2} \sum_{p,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(p)},
  \label{eq:H_int_perm_start}
\end{align}
where the first sum,
\begin{align}
  \sum_{\substack{p<q\\\mu,\nu}} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  = \sum_{\substack{p<q\\\mu,\nu}} \op{\mu}{\nu}^{(p)} \op{\nu}{\mu}^{(q)}
  = \sum_{\substack{p<q\\\mu,\nu}} \op{\mu\nu}{\nu\mu}^{(p,q)}
  = \sum_{p<q} \Pi_{pq}
  \label{eq:spin_perm}
\end{align}
is simply a sum over all permutations $\Pi_{pq}$ of subsystems $p$ and
$q$; and
\begin{align}
  \sum_{p,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(p)}
  = \sum_{p,\mu,\nu} \op{\mu}{\nu}^{(p)} \op{\nu}{\mu}^{(p)}
  = n \sum_{p,\mu} \op{\mu}{\mu}^{(p)}
  = n \sum_p \1^{(p)}
  = n N
  \label{eq:spin_const}
\end{align}
is merely a constant.  Substituting \eqref{eq:spin_perm} and
\eqref{eq:spin_const} into \eqref{eq:H_int_perm_start} yields
\begin{align}
  H_{\t{int}} = -u \sum_{p<q} \Pi_{pq} - \f12 N n u
  \simeq -u\sum_{p<q} \Pi_{pq},
  \label{eq:H_int_perm}
\end{align}
where $\simeq$ denotes equality up to an overall constant with no
physical consequence.  The form of the interaction Hamiltonian
$H_{\t{int}}$ makes it evident that its ground-state manifold
$\M_{\FS}$ is the set of {\it fully symmetric} (FS) states that have
eigenvalue 1 with respect to all pairwise permutation operators
$\Pi_{pq}$, and thereby also all $\Pi_\sigma\in\Pi$.  The fully
symmetric manifold $\M_\FS$ is spanned by a basis $\set{\ket{\v m}}$
of fully symmetric states that can be uniquely identified by a list of
integers $\v m\equiv\p{m_0,m_1,\cdots,m_{n-1}}\in\ZZ_{N+1}^n$, where
$m_s$ indicates the total occupation number of subsystem state $s$,
and $\sum_sm_s=N$.  In the case of $n=2$, the fully symmetric manifold
$\M_\FS$ is precisely the Dicke manifold\cite{dicke1954coherence} of
$N$ qubits, spanned by Dicke states $\set{\ket{m_0,m_1}}$ for which
$m_0+m_1=N$ and the integer $m_0$ ($m_1$) indicates the total
occupation number of qubit state $\ket{0}$ ($\ket{1}$).  The dimension
of fully symmetric manifold $\M_\FS$ is determined by the number of
ways to assign $N$ subsystems to $n$ states, which is
${N+n-1 \choose n-1}$, and the energy of a state
$\ket{\psi_\FS}\in\M_\FS$ with respect to the interaction Hamiltonian
$H_{\t{int}}$ in \eqref{eq:H_int_perm} is
$-u{N \choose 2}=-\p{1/4}N\p{N-1}u$.

@@

The single-particle Hamiltonian $H_{\t{lat}}^{(\phi)}$ breaks
permutational symmetry, thereby coupling states within the fully
symmetric manifold $\M_\FS$ to states outsides it.  If SOC is
sufficiently weak, however, transitions outside the fully symmetric
manifold $\M_\FS$ are energetically suppressed by interactions, and we
can account for the effect of SOC perturbatively.  In the regime of
weak SOC with $\abs{\phi}\ll1$, by considering
\begin{enumerate*}
\item all $L\le12$ for arbitrary $n$, and
\item all $L<n$ for all $n\le25$
\end{enumerate*}
(and conjecturing the same functional form for all $L$ and $n$), we
find that
% expand Clebsch-Gordan coefficients in a power series according to
% \url{functions.wolfram.com/07.38.06.0008.01}
\begin{align}
  A_L^{(\phi)}
  = \p{-1}^{L+\lfloor L/2\rfloor} \xi_L \phi^L + O\p{\phi^{L+2}},
  \label{eq:A_L_phi_small}
\end{align}
where $\xi_L$ is a scale factor defined in \eqref{eq:scale_fac}.  A
perturbative treatment of SOC through second order in the SOC angle
$\phi$ yields the following effective Hamiltonian in the fully
symmetric manifold $\M_\FS$ (see Appendix \ref{sec:SOC_pert}):
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
\end{align}
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_p\sp{\sin\p{pa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi}
\end{align}
where
\begin{align}
  \EE_p\sp{X_p} \equiv \f1N \sum_p X_p,
  &&
  \var_p\sp{X_p} \equiv \EE_p\sp{\p{X_p-\EE_q\sp{X_q}}}
  = \EE_p\sp{X_p^2}-\EE_q\sp{X_q}^2,
\end{align}
are the mean and variance of $X_p$ over all occupied quasi-momenta
$p$.  In the case of SU(2), the effective Hamiltonians in
\eqref{eq:H_eff_1_phi} and \eqref{eq:H_eff_2_phi} reduce to the
spin-squeezing Hamiltonians derived in Ref.~\cite{he2019engineering}.
To facilitate comparison, we note that in the case of SU(2) the drive
operator $\D_{1,0}$ is precisely the collective spin-$z$ operator
$S_\z$ that appears in Ref.~\cite{he2019engineering}.  Furthermore,
drive operators $D_{LM}$ vanish when $L\ge n$, so in the case of SU(2)
the operator $\D_{2,0}$ has no contribution to the effective
Hamiltonians in \eqref{eq:H_eff_2_phi} and \eqref{eq:H_eff_2_phi}.

@@

adjacent lattice sites; we have added an overall factor of $\xi_1$ to
simplify later expressions; $\op{r}{s}$ for $r,s\in\set{\up,\dn}$ is a

@@

operator.  If the detuning is large, $\Delta\gg\Omega_m$, then at
second order in perturbation theory the effective action of

@@

where $H_{\t{Raman},\tau}^{(\phi)}$ generates a nuclear spin
transition $\mu\to\mu\pm\tau$:

@@

  \Omega_{1,1} &\propto \Omega_0 \p{\Omega_+ + \Omega_-}, \\

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future directions}

\begin{itemize}
\item Visualization with generalized Bloch spheres
  \begin{itemize}
  \item Single-particle eigenstates of drive operators
  \item Single-particle dynamics induced by drive operators
  \end{itemize}
\item Classifying collective states (e.g.~eigenstates of interactions,
  spin-wave-like states)
\item Implications of SOC
  \begin{itemize}
  \item Special angles $\phi$
  \end{itemize}
\item Generalizations of squeezing beyond SU(2)?
\item Consider different interactions
  \begin{itemize}
  \item Include electronic states
  \item Super-exchange regime
  \item Non-uniform (``disordered'') interactions
  \end{itemize}
\item Consider different drives
\end{itemize}

@@

The effective Hamiltonians in \eqref{eq:H_eff_1} and
\eqref{eq:H_eff_2} are respectively first and second order in the SOC
Hamiltonian $H_{\t{lat}}^{(\phi)}$.  Here we extract the leading and
next-to-leading order dependence of these effective Hamiltonians on
the SOC angle $\phi$ using the expansion
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi_apndx} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_p\sp{\sin\p{pa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi_apndx}
\end{align}
where the factors $\xi_L$ are defined in \eqref{eq:scale_fac},
\begin{align}
  \var_q\sp{X_q} \equiv \cov_{p,q}\sp{X_p,X_q}
  = \EE_p\sp{\p{X_p-\EE_q\sp{X_q}}^2}
\end{align}
is the variance of $X_q$ over all occupied quasi-momenta $q$, and we
have neglected scalar terms $\sim\D_{0,0}\propto\1$ that have no
physical consequence.  Note that in the case of SU(2), the drive
operator $\tilde\D_{1,0}$ is precisely the collective spin-$z$
operator $S_\z$ that appears in Ref.~\cite{he2019engineering}.
Furthermore, drive operators $D_{LM}$ vanish when $L\ge n$, so in the
case of SU(2) the operator $\tilde\D_{2,0}$ has no contribution to the
effective Hamiltonians in \eqref{eq:H_eff_2_phi_apndx} and
\eqref{eq:H_eff_2_phi_apndx}.

>>>>

For our final re-expression of collective SU($n$)-symmetric
interactions, we abandon the use of single-spin operators entirely to
express the interaction Hamiltonian $H_{\t{int}}$ in terms of
operators that permute tensor factors (subsystems) of a Hilbert space
with an $N$-fold tensor product structure (e.g.~an array of nuclear
spins).  Specifically, we denote the Hilbert space of an $n$-level
system by $\H_n$, denote its $N$-fold tensor product by
$\H_n^{\otimes N}$, and identify the set $\Pi\equiv\set{\Pi_\sigma}$
of operators that permute tensor factors of $\H_n^{\otimes N}$
according to elements $\sigma\in\SS_N$ of the symmetric group of order
$N$.  We then expand
\begin{align}
  H_{\t{int}}
  = -\f{u}{2} \sum_{p,q,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  = -u\sum_{\substack{p<q\\\mu,\nu}} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  - \f{u}{2} \sum_{p,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(p)},
  \label{eq:H_int_perm_start}
\end{align}
where the first sum,
\begin{align}
  \sum_{\substack{p<q\\\mu,\nu}} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  = \sum_{\substack{p<q\\\mu,\nu}} \op{\mu}{\nu}^{(p)} \op{\nu}{\mu}^{(q)}
  = \sum_{\substack{p<q\\\mu,\nu}} \op{\mu\nu}{\nu\mu}^{(p,q)}
  = \sum_{p<q} \Pi_{pq}
  \label{eq:spin_perm}
\end{align}
is simply a sum over all permutations $\Pi_{pq}$ of subsystems $p$ and
$q$; and the second sum,
\begin{align}
  \sum_{p,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(p)}
  = \sum_{p,\mu,\nu} \op{\mu}{\nu}^{(p)} \op{\nu}{\mu}^{(p)}
  = n \sum_{p,\mu} \op{\mu}{\mu}^{(p)}
  = n \sum_p \1^{(p)}
  = n N,
  \label{eq:spin_const}
\end{align}
is merely a constant.  Substituting \eqref{eq:spin_perm} and
\eqref{eq:spin_const} into \eqref{eq:H_int_perm_start} yields
\begin{align}
  H_{\t{int}} = -u \sum_{p<q} \Pi_{pq} - \f12 N n u
  \simeq -u\sum_{p<q} \Pi_{pq},
  \label{eq:H_int_perm}
\end{align}
where $\simeq$ denotes equality up to an overall constant with no
physical consequence.  The form of the interaction Hamiltonian
$H_{\t{int}}$ makes it evident that its ground-state manifold
$\M_{\FS}$ is the set of {\it fully symmetric} (FS) states that have
eigenvalue 1 with respect to all pairwise permutation operators
$\Pi_{pq}$.  The fully symmetric manifold $\M_\FS$ is spanned by a
basis $\set{\ket{\v m}}$ of fully symmetric states that can be
uniquely identified by a list of integers
$\v m\equiv\p{m_0,m_1,\cdots,m_{n-1}}\in\ZZ_{N+1}^n$, where $m_s$
indicates the total occupation number of subsystem state $s$, and
$\sum_sm_s=N$.  In the case of $n=2$, the fully symmetric manifold
$\M_\FS$ is precisely the Dicke manifold\cite{dicke1954coherence} of
$N$ qubits, spanned by Dicke states $\set{\ket{m_0,m_1}}$ for which
$m_0+m_1=N$ and the integer $m_0$ ($m_1$) indicates the total
occupation number of qubit state $\ket{0}$ ($\ket{1}$).  The dimension
of fully symmetric manifold $\M_\FS$ is determined by the number of
ways to assign $N$ subsystems to $n$ states, which is
${N+n-1 \choose n-1}$, and the energy of any fully symmetric state
$\ket{\psi_\FS}\in\M_\FS$ with respect to the interaction Hamiltonian
$H_{\t{int}}$ in \eqref{eq:H_int_perm} is
$-u{N \choose 2}=-\p{1/4}N\p{N-1}u$.

@@

The single-particle Hamiltonian $H_{\t{lat}}^{(\phi)}$ breaks
permutational symmetry, thereby coupling states within the fully
symmetric manifold $\M_\FS$ to states outsides it.  If SOC is
sufficiently weak, however, transitions outside the fully symmetric
manifold $\M_\FS$ are energetically suppressed by interactions, and we
can account for the effect of SOC perturbatively.  In the regime of
weak SOC with $\abs{\phi}\ll1$, by considering
\begin{enumerate*}
\item all $L\le12$ for arbitrary $n$, and
\item all $L<n$ for all $n\le25$
\end{enumerate*}
(and conjecturing the same functional form for all $L$ and $n$), we
find that
% expand Clebsch-Gordan coefficients in a power series according to
% \url{functions.wolfram.com/07.38.06.0008.01}
\begin{align}
  A_L^{(\phi)}
  = \p{-1}^{L+\lfloor L/2\rfloor} \xi_L \phi^L + O\p{\phi^{L+2}},
  \label{eq:A_L_phi_small}
\end{align}
where $\xi_L$ is a scale factor defined in \eqref{eq:scale_fac}.  A
perturbative treatment of SOC through second order in the SOC angle
$\phi$ yields the following effective Hamiltonian in the fully
symmetric manifold $\M_\FS$ (see Appendix \ref{sec:SOC_pert}):
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
\end{align}
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_q\sp{\sin\p{qa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi}
\end{align}
where
\begin{align}
  \EE_q\sp{X_q} \equiv \f1N \sum_q X_q,
  &&
  \var_q\sp{X_q} \equiv \EE_p\sp{\p{X_p-\EE_q\sp{X_q}}}
  = \EE_p\sp{X_p^2}-\EE_q\sp{X_q}^2,
\end{align}
are the mean and variance of $X_q$ over all occupied quasi-momenta
$q$.  In the case of SU(2), the effective Hamiltonians in
\eqref{eq:H_eff_1_phi} and \eqref{eq:H_eff_2_phi} reduce to the
spin-squeezing Hamiltonians derived in Ref.~\cite{he2019engineering}.
To facilitate comparison, we note that in the case of SU(2) the drive
operator $\D_{1,0}$ is precisely the collective spin-$z$ operator
$S_\z$ that appears in Ref.~\cite{he2019engineering}.  Furthermore,
drive operators $D_{LM}$ vanish when $L\ge n$, so in the case of SU(2)
the operator $\D_{2,0}$ has no contribution to the effective
Hamiltonians in \eqref{eq:H_eff_2_phi} and \eqref{eq:H_eff_2_phi}.

@@

adjacent lattice sites; $\op{r}{s}$ for $r,s\in\set{\up,\dn}$ is a

@@

operator.  If the detuning is large, $\abs{\Delta}\gg\abs{\Omega_m}$,
then at second order in perturbation theory the effective action of

@@

where $H_{\t{Raman},\tau}^{(\phi)}$ generates nuclear spin transitions
$\mu\to\mu\pm\tau$:

@@

@@

  \Omega_{1,1} &\propto \Omega_0 \p{\Omega_+ + \Omega_-},

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Possible future directions}

\begin{itemize}
\item Visualization of multilevel states
  \begin{itemize}
  \item Single-particle eigenstates of drive operators
  \item Single-particle dynamics induced by drive operators
  \item Collective states and dynamics
  \end{itemize}
\item Classifying eigenstates of SU($n$)-symmetric interactions
\item Implications of SOC
  \begin{itemize}
  \item Special angles $\phi$
  \end{itemize}
\item Generalizations of squeezing beyond SU(2)?
\item Consider different interactions
  \begin{itemize}
  \item Include electronic states
  \item Super-exchange regime
  \item Non-uniform (``disordered'') interactions
  \end{itemize}
\item Consider different drives
\end{itemize}

@@

The effective Hamiltonians in \eqref{eq:H_eff_1} and
\eqref{eq:H_eff_2} are respectively first and second order in the SOC
Hamiltonian $H_{\t{lat}}^{(\phi)}$.  Here we extract the leading and
next-to-leading order dependence of these effective Hamiltonians on
the SOC angle $\phi$ using the expansion
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi_apndx} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_q\sp{\sin\p{qa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi_apndx}
\end{align}
where the factors $\xi_L$ are defined in \eqref{eq:scale_fac},
\begin{align}
  \var_q\sp{X_q} \equiv \cov_{p,q}\sp{X_p,X_q}
  = \EE_p\sp{\p{X_p-\EE_q\sp{X_q}}^2}
  = \EE_p\sp{X_p^2} - \EE_q\sp{X_q}^2
\end{align}
is the variance of $X_q$ over all occupied quasi-momenta $q$, and we
have neglected scalar terms $\sim\D_{0,0}\propto\1$ that have no
physical consequence.  Note that in the case of SU(2), the drive
operator $\tilde\D_{1,0}$ is precisely the collective spin-$z$
operator $S_\z$ that appears in Ref.~\cite{he2019engineering}.
Furthermore, drive operators $D_{LM}$ vanish when $L\ge n$, so in the
case of SU(2) the operator $\tilde\D_{2,0}$ has no contribution to the
effective Hamiltonians in \eqref{eq:H_eff_2_phi_apndx} and
\eqref{eq:H_eff_2_phi_apndx}.
